---
title: "Wheelfun_Rentals_Weather_Analysis"
author: "Robert Mounce"
date: "2024-11-28"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Wheel Fun Rentals Analysis

## Objective of Analysis

The purpose of this analysis is to explore the relationships between
park attendance at Orange County parks, bicycle repair reports, and
weather conditions over a defined time period. By analyzing these
variables, the study aims to identify patterns and correlations that can
provide insights into operational planning, resource allocation, and
potential impacts of weather on park and equipment usage. Data Sources
and Details

1.  Weather Data • Description: Daily weather records from December 31,
    2018, to January 30, 2022, covering Orange County. • Variables:
    Rainfall, wind speed, and other weather conditions. • Source:
    National Oceanic and Atmospheric Administration (NOAA). • Dataset on
    Kaggle
    <https://www.kaggle.com/datasets/robertmounce/oc-weather-data/data>

2.  Park Attendance Data • Description: Monthly attendance records for
    Yorba Linda Regional Park and Mile Square Park from January 2019 to
    December 2021. • Purpose: Provides insights into how weather
    conditions may influence visitor numbers. • Source: Orange County
    Parks Data on Kaggle.
    <https://www.kaggle.com/datasets/robertmounce/orange-county-parks-data/data>

3.  Mechanic Log Data • Description: Detailed repair reports from 2019
    to 2020, collected through a system integrating an API and QR codes
    onsite. • Purpose: Tracks bicycle and equipment repair trends in
    relation to park attendance and weather patterns. • Data Cleaning
    and Analysis: Notebook on
    Kaggle.<https://www.kaggle.com/code/robertmounce/wheel-fun-rentals-cleaning-and-analysis/notebook>

```{r Libraries and Datasets}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(lmtest)
library(sf)
library(sandwich)
library(leaflet)
library(lubridate)
library(car)

# Load the data sets
weather_data <- read.csv("cleaned_weather_data.csv")
attendance_data <- read.csv("Park_data.csv")
repairs_data <- read.csv("Repair_data.csv")

# Standardized column names for consistency 
colnames(attendance_data)[colnames(attendance_data) == "Park"] <- "Location"
colnames(weather_data)[colnames(weather_data) == "NAME"] <- "Location"
colnames(attendance_data)[colnames(attendance_data) == "Mon.Year"] <- "Date"
colnames(attendance_data)[colnames(attendance_data) == "Park"] <- "Location"
colnames(attendance_data)[colnames(attendance_data) == "Visitors.Month"] <- "Month_Attendance"
colnames(attendance_data)[colnames(attendance_data) == "Visitors.day"] <- "Average_Visitors_Per_Day"

# Drop unneeded columns
attendance_data <- attendance_data %>%
  select(-c(Month..num., Year, Month))

# Standardize date columns
attendance_data$Date <- as.Date(attendance_data$Date, format = "%Y-%m-%d") #Monthly recorded
weather_data$Date <- as.Date(weather_data$Date)  # Daily recorded
repairs_data$Down_Date <- as.Date(repairs_data$Down_Date) # Daily recorded 
repairs_data$Fixed_Date <- as.Date(repairs_data$Fixed_Date) # Daily recorded
```

## Filter and Aggregate Weather Data

1.  Create a map to see what stations we want to have for our locations.

```{r Weather Stations}
# How close are these stations to our Locations
parks <- data.frame(
  Name = c("Yorba Linda", "Mile Square", "Huntington Beach"),
  Latitude = c(33.8714, 33.7190, 33.653351),
  Longitude = c(-117.7638, -117.9383, -117.994209)
)
# Where are the stations are at
weather_stations <- weather_data %>%
  select(Location, LATITUDE, LONGITUDE) %>%
  distinct()

# Create an interactive map
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    # Weather stations
    data = weather_stations,
    lat = ~LATITUDE, lng = ~LONGITUDE,
    popup = ~Location,
    color = "red",
    radius = 5,
    label = ~Location
  ) %>%
  # Parks
  addCircleMarkers(
    data = parks,
    lat = ~Latitude, lng = ~Longitude,
    popup = ~Name,
    color = "blue",
    radius = 6,
    label = ~Name
  ) %>%
  addLegend(
    position = "topright",
    colors = c("blue", "red"),
    labels = c("Parks", "Weather Stations"),
    title = "Locations"
  )
```

### Weather Data Prep

Based on the analysis of the data and geographic proximity, I have
chosen to use only two weather stations for rain data. The Huntington
Beach 2.0 ESE station was selected for both the Waterfront and Mile
Square Park due to its proximity to these locations. For wind speed, I
opted to use data from nearby airports rather than weather stations.
This decision was based on observed inconsistencies and unreliability in
wind speed measurements at the weather stations. - Waterfront: - Rain
Data: HUNTINGTON BEACH 2.0 ESE, CA US - Mile Square Park: - Rain Data:
HUNTINGTON BEACH 2.0 ESE, CA US - Wind Speed: SANTA ANA JOHN WAYNE
AIRPORT, CA US (Swan Boats) - Yorba Linda Regional Park: - Rain Data:
ANAHEIM 7.3 E, CA US - Wind Speed: FULLERTON MUNICIPAL AIRPORT, CA US
(Swan Boats)

Wind speed data is particularly relevant for swan boat operations at
Mile Square Park and Yorba Linda Regional Park. Swan boat rentals are
suspended when wind speeds exceed 7 mph, as higher winds can make
boating unsafe and hard for customers to ride. Additionally, we
hypothesize that strong winds may contribute to boat damage.
Specifically, late-night winds could loosen the boats from the bouy they
are attached to, causing them to hit the pond edges and sustain holes.

```{r Wind Data Selection}
# Add AvgTemp column for average temperature
weather_data <- weather_data %>%
  mutate(AvgTemp = (MaxTemp + MinTemp) / 2 )

# Filter relevant weather station
filtered_weather_data <- weather_data %>%
  filter(Location %in% c("HUNTINGTON BEACH 2.0 ESE, CA US", "ANAHEIM 7.3 E, CA US"))

# Separate weather by location
yorba_weather <- filtered_weather_data%>%
  filter(Location == "ANAHEIM 7.3 E, CA US") %>%
  select(Date,Rain, Precipitation)

mile_weather <- filtered_weather_data%>%
  filter(Location == "HUNTINGTON BEACH 2.0 ESE, CA US") %>% 
  select(Date, Rain, Precipitation)
# Combine weather data with wind speed from airport stations
mile_weather <- merge(mile_weather, weather_data %>%
                        filter(Location == "SANTA ANA JOHN WAYNE AIRPORT, CA US")%>% 
                        select(Date, WindSpeed, AvgTemp), by = "Date", all.x = TRUE)


yorba_weather <- merge(yorba_weather, weather_data %>%
                        filter(Location == "FULLERTON MUNICIPAL AIRPORT, CA US")%>% 
                        select(Date, WindSpeed, AvgTemp), by = "Date", all.x = TRUE)

```

## Repair Data Preparation

-   filtering out the repair data from the locations and getting the
    number of reports grouped by the date for analysis.

```{r Clean and Repair reports}
# Drop these columns Mechanic, Down_Date_ordinal, Category, Status, Mechanic_Notes
repairs_data <- repairs_data%>%
  select(-matches("Mechanic|Down_Date_ordinal|Category|Status| Mechanic_Notes|PartComment"))

# Filter the repair data by the Locations
mile_repair <- repairs_data %>%
  filter(Location %in% c("Warner", "Edinger")) %>%
  group_by(Down_Date) %>%
  summarise(Reports = n())

yorba_repair <- repairs_data%>%
  filter(Location == "Yorba") %>%
  group_by(Down_Date) %>%
  summarise(Reports = n())

waterfront_repair <- repairs_data %>%
  filter(Location == "Waterfront")
```

Attendance Data

```{r}
# Convert the attendance data column to a Month Column
attendance_data <- attendance_data %>%
  mutate(Month = floor_date(Date, "month")) %>%  # Extract year and month
  select(Month, Month_Attendance)  # Keep only the relevant columns
```

```{r}
# Add Month Column to daily data for merging
mile_df <- mile_repair%>%
  mutate(Month = floor_date(Down_Date, "month"))

# Prepare repair data with a month column 
yorba_df <- yorba_repair %>% 
  mutate(Month = floor_date(Down_Date, "month"))

# Merge Attendance 
mile_df <- mile_df %>% 
  left_join(attendance_data, by = "Month")

yorba_df <- yorba_df%>% 
  left_join(attendance_data, by = "Month")
# Merge weather
yorba_df <- yorba_df %>%
  left_join(yorba_weather, by = c("Down_Date" = "Date"))

mile_df<- mile_df%>%
  left_join(mile_weather, by = c("Down_Date" = "Date"))

```

```{r}
# Month column 
mile_df <- mile_df %>% 
  mutate(Month = floor_date(Down_Date, "month"))
yorba_df <- yorba_df %>% 
  mutate(Month = floor_date(Down_Date, "month"))

# Group and sumarize
mile_df <- mile_df %>%
  group_by(Month) %>%
  summarise(
    Total_Reports = sum(Reports, na.rm = TRUE),
    Avg_WindSpeed = mean(WindSpeed, na.rm = TRUE),
    Total_Precipitation = sum(Precipitation, na.rm = TRUE),
    Avg_Temperature = mean(AvgTemp, na.rm = TRUE),
    Attendance = sum(Month_Attendance, na.rm = TRUE)
  )

yorba_df <- yorba_df %>%
  group_by(Month) %>%
  summarise(
    Total_Reports = sum(Reports, na.rm = TRUE),
    Avg_WindSpeed = mean(WindSpeed, na.rm = TRUE),
    Total_Precipitation = sum(Precipitation, na.rm = TRUE),
    Avg_Temperature = mean(AvgTemp, na.rm = TRUE),
    Attendance = sum(Month_Attendance, na.rm = TRUE)
  )

```

# Linear Regression Analysis

## Yorba Linda Linear Regression

## 1. Objective To explore the relationship between repair reports at

```         
Yorba Regional Park (Total_Reports) and weather conditions
(Avg_WindSpeed, Total_Precipitation), as well as park
attendance(Attendance) Dependent Variable: Total_Reports Independent
Variables:
```

-   Avg_WindSpeed: Average daily windspeed (mph)
-   Total_Precipitation: Total daily precipitation (inches)
-   Attendance: Monthly park attendance

## 2. Key Findings

### 1. Attendance

-   Statistically significant (p = 0.000589)
-   Positive relationship with Total_Reports: A higher number of park
    vistors is strongly associated with repair reports

### 2. Weather Variables

-   Avg_WindSpeed and Total_Precipitation remain statistically
    significant, suggesting they have little direct impact on
    Total_Reports

## Conclusion:

Attendance is the key driver of Total_reports suggesting that
operational planning should align with park attendance levels Weather
conditions do not have a direct, significant impact but could contribute
indirectly when combined with attendance

```{r Yorba Linda Model Analysis}
yorba_model <- lm(Total_Reports ~ Avg_WindSpeed + Total_Precipitation + Attendance, data = yorba_df)
summary(yorba_model)
```

```{r Yorba Residuals}
# Plot residuals 
par(mfrow = c(2,2))
# Histogram of residuals
hist(residuals(yorba_model), main = "Histogram of Residuals" , xlab = "Residuals", breaks = 20)
# Heteroskedastcity
bptest(yorba_model)
```

```{r}
yorba_df$Residuals <- residuals(yorba_model)
outlier_data <- yorba_df %>%
  filter(Residuals > 10 | Residuals < -10)
print(outlier_data)
```

```{r Yorba multicollinearity}
vif(yorba_model)
```

```{r}
yorba_att_weather_model <- lm(Total_Reports ~ Attendance * Total_Precipitation, data = yorba_df)
summary(yorba_att_weather_model)
```

```{r Visualize Yorba Attendance}
# Attendance vs Total_Reports
ggplot(yorba_df, aes(x = Attendance, y = Total_Reports)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Attendance vs Total_Reports", x = "Attendance", y = "Total Reports")
```
